### 2025-01-19 paper/icml25/first/sects/background.tex
prefer Cref on ref

To solve an MDP $\mathcal{M}$, SHAC trains a policy $\pi_\theta$ by minimizing the loss function $\mathcal{L}_\theta^{\pi}:\mathcal{T}^N\rightarrow\mathbb{R}$ for a batch of $N$ trajectories $B = \{\tau_1, \ldots, \tau_N\}$ sampled from the MDP. The loss is defined as:  
\begin{equation}\label{eq:shac}
    \mathcal{L}_\theta^{\pi}(B) = \frac{1}{NH}\sum_{\tau\in B} \sum_{t=t_0}^{t_0+H-1} \gamma^{t-t_0} R(s_t^\tau, a_t^\tau) + \gamma^H V_\theta(s^\tau_{t_0+T})
\end{equation}
where $N$ is the number of sampled trajectories, $H \ll T$ is a time horizon, $s_t^\tau$ is the state at time $t$ for trajectory $\tau$, $a_t^\tau$ is the action taken at time $t$ for trajectory $\tau$, and $V(s): \mathcal{S} \rightarrow \mathbb{R}$ is the value function estimating the expected return starting from state $s$. From Equation~\ref{eq:shac}, it is apparent that SHAC assumes that the policy, reward, value, and transition functions are all differentiable, i.e., $\pi_\theta,V_\theta,F,R \in \mathcal{C}^1$. While this is an common requirement for policy and value functions, differentiable transition and reward functions are rarely available.

@@

Finally, Figure~\ref{fig:shac} presents an unrolled view of the SHAC algorithm.

>>>>

To solve an MDP $\mathcal{M}$, SHAC trains a policy $\pi_\theta$ by minimizing the loss function $\mathcal{L}_\theta^{\pi}:\mathcal{T}^N\rightarrow\mathbb{R}$ for a batch of $N$ trajectories $B = \{\tau_1, \ldots, \tau_N\}$ sampled from the MDP. The loss is defined as:  
\begin{equation}\label{eq:shac}
    \mathcal{L}_\theta^{\pi}(B) = \frac{1}{NH}\sum_{\tau\in B} \sum_{t=t_0}^{t_0+H-1} \gamma^{t-t_0} R(s_t^\tau, a_t^\tau) + \gamma^H V_\theta(s^\tau_{t_0+T})
\end{equation}
where $N$ is the number of sampled trajectories, $H \ll T$ is a time horizon, $s_t^\tau$ is the state at time $t$ for trajectory $\tau$, $a_t^\tau$ is the action taken at time $t$ for trajectory $\tau$, and $V(s): \mathcal{S} \rightarrow \mathbb{R}$ is the value function estimating the expected return starting from state $s$. From \Cref{eq:shac}, it is apparent that SHAC assumes that the policy, reward, value, and transition functions are all differentiable, i.e., $\pi_\theta,V_\theta,F,R \in \mathcal{C}^1$. While this is an common requirement for policy and value functions, differentiable transition and reward functions are rarely available.

@@

Finally, \Cref{fig:shac} presents an unrolled view of the SHAC algorithm.

### 2025-01-19 paper/icml25/first/sects/discussion.tex
prefer Cref on ref

Note that \fname{} does not aim to outperform SHAC, it merely aims to perform comparably with SHAC while being applicable even in non-differentiable environment. The result shown in Sect.~\ref{sect:experiments} seems to support this conclusion. More specifically, let us answer the research questions posed in Sect.~\ref{sect:introduction}.

>>>>

Note that \fname{} does not aim to outperform SHAC, it merely aims to perform comparably with SHAC while being applicable even in non-differentiable environment. The result shown in \Cref{sect:experiments} seems to support this conclusion. More specifically, let us answer the research questions posed in \Cref{sect:introduction}.

### 2025-01-19 paper/icml25/first/sects/experiments.tex
fixed main-transformer.pdf
prefer Cref on ref

\begin{figure*}[!t]
    \centering
    \begin{subfigure}[b]{0.30\textwidth}
        \includegraphics[width=\textwidth]{figs/dispersion-1-mlp.pdf}
        \caption{Dispersion, MLP, 1 agent}
        \label{fig:dispersion-mlp-1}
    \end{subfigure}
    \begin{subfigure}[b]{0.30\textwidth}
        \includegraphics[width=\textwidth]{figs/dispersion-3-transformer.pdf}
        \caption{Dispersion, Transformer, 3 agents}
        \label{fig:dispersion-transformer-3}
    \end{subfigure}
    \begin{subfigure}[b]{0.30\textwidth}
        \includegraphics[width=\textwidth]{figs/dispersion-5-transformer.pdf}
        \caption{Dispersion, Transformer, 5 agents}
        \label{fig:dispersion-transformer-5}
    \end{subfigure}

    \begin{subfigure}[b]{0.30\textwidth}
        \includegraphics[width=\textwidth]{figs/transport-1-mlp.pdf}
        \caption{Transport, MLP, 1 agent}
        \label{fig:transport-mlp-1}
    \end{subfigure}
    \begin{subfigure}[b]{0.30\textwidth}
        \includegraphics[width=\textwidth]{figs/transport-3-transformer.pdf}
        \caption{Transport, Transformer, 3 agents}
        \label{fig:transport-transformer-3}
    \end{subfigure}
    \begin{subfigure}[b]{0.30\textwidth}
        \includegraphics[width=\textwidth]{figs/transport-5-transformer.pdf}
        \caption{Transport, Transformer, 5 agents}
        \label{fig:transport-transformer-5}
    \end{subfigure}

    \begin{subfigure}[b]{0.30\textwidth}
        \includegraphics[width=\textwidth]{figs/discovery-1-mlp.pdf}
        \caption{Discovery, MLP, 1 agent}
        \label{fig:discovery-mlp-1}
    \end{subfigure}
    \begin{subfigure}[b]{0.30\textwidth}
        \includegraphics[width=\textwidth]{figs/discovery-3-transformer.pdf}
        \caption{Discovery, Transformer, 3 agents}
        \label{fig:discovery-transformer-3}
    \end{subfigure}
    \begin{subfigure}[b]{0.30\textwidth}
        \includegraphics[width=\textwidth]{figs/discovery-5-transformer.pdf}
        \caption{Discovery, Transformer, 5 agents}
        \label{fig:discovery-transformer-5}
    \end{subfigure}

    \begin{subfigure}[b]{0.30\textwidth}
        \includegraphics[width=\textwidth]{figs/sampling-1-mlp.pdf}
        \caption{Discovery, MLP, 1 agent}
        \label{fig:sampling-mlp-1}
    \end{subfigure}
    \begin{subfigure}[b]{0.30\textwidth}
        \includegraphics[width=\textwidth]{figs/sampling-3-transformer.pdf}
        \caption{Discovery, Transformer, 3 agents}
        \label{fig:sampling-transformer-3}
    \end{subfigure}
    \begin{subfigure}[b]{0.30\textwidth}
        \includegraphics[width=\textwidth]{figs/sampling-5-transformer.pdf}
        \caption{Discovery, Transformer, 5 agents}
        \label{fig:sampling-transformer-5}
    \end{subfigure}

    \caption{Comparison between \fname{}, PPO, and SHAC for increasing number of agents for Dispersion, Transport, Discovery, and Sampling scenarios for the Transformer Architecture.}
    \label{fig:experiments}
\end{figure*}

@@

\paragraph{Scenarios}
As mentioned previously, we choose a few multi-agent scenarios from the VMAS\footnote{\url{https://github.com/proroklab/VectorizedMultiAgentSimulator}} package to test \fname{} aganist SHAC and PPO. In particular, we choose the following scenario:
\begin{compactitem}
    \item \textbf{Dispersion} (Fig.~\ref{fig:dispersion}): 
        \sentence{Environment} a scenario where $n$ agents need to reach $n$ goal locations, in such case the agent receives a sparse reward of $1$. Agents observe only their own position, velocity and relative position to each goal. 
        \sentence{Differentiability} Despite the transition function being differentiable, the reward function is not. Therefore, only PPO and \fname{} are applicable to this scenario. 
        \sentence{World Model} The world model has enough information to predict the next states accurately.
        \sentence{Max Reward} The maximum reward is $n$.
    \item \textbf{Transport} (Fig.~\ref{fig:transport}): 
        \sentence{Environment} a scenario where $n$ agents need to collaborate to push a package into a goal location. Agents observe their own position, velocity, relative position to the package and relative position between package and the goal. The agents receive a reward proportional to the distance between package and the goal. 
        \sentence{Differentiability} In this scenario, both the reward and transition functions are differentiable. 
        \sentence{World Model} The world model has enough information to predict the next states accurately.
        \sentence{Max Reward} The maximum reward varies from environement to environment. It is the scaled distance between the package and the goal times the number of agents.
    \item \textbf{Discovery} (Fig.~\ref{fig:discovery}): 
        \sentence{Environment} a scenario where $n$ agents need to collect $k$ goal place randomly. A goal is collected if at least $s$ agents stand in proximity to it, in this case they receive a reward of $1$. If agents collide with each other, they receive a penalty. Every agent observes its position, velocity, lidar measurements to other agents and goals. When $n=1$ we set $s=1$ and $k=7$. When $n>1$ we set $s=2$.
        \sentence{Differentiability} In this scenario, only the transition function is differentiable. Therefore, only PPO and \fname{} are applicable to this scenario. 
        \sentence{Max Reward} The maximum rewards depends on the number evaluation steps, but for early stopping purposes, we set it to $k$. 
        \sentence{World Model} Note that, a world model (or an action world model) that uses only the agents observation cannot predict the evolution correctly, as goals may not be visible to the agents. For this reason, this scenario is particularly challenging for \fname{} as we do not provide the full world state to the world model.
     \item \textbf{Sampling} (Fig.~\ref{fig:sampling}): 
        \sentence{Environment} a scenario where $n$ agents need to collect rewards from a grid. The reward are sampled from $k$ gaussians. Agents observe their own position, velocity, lidar values to each other, and rewards in a 3x3 grid around themselves.  
        \sentence{Differentiability} In this scenario, both the reward and transition functions are differentiable. 
        \sentence{Max Reward} The maximum reward depends on where the gaussians are placed.
        \sentence{World Model} Similarly to discovery, this scenario is particularly challenging for \fname{} as we do not provide the full world state, but only the agents observations, to the world model.
\end{compactitem}

@@

\paragraph{Hyperparameters}
For all networks we use learning rate of $1e\text{-}3$ with Adam Optimizer~\cite{Kingma14}. For all scenarios, each training episode consists of $512$ environments of $32$ steps each. Each validation episode is composed of $512$ environments of $512$ steps each. We employ early stopping when the agents achieve $90\%$ of the maximum reward in $90\%$ of the episode's environments. If early stopping is not triggered, we stop training after $20,000$ episodes. The discount factor is set to $0.99$ and lambda factor is set to $0.95$. We report results for increasing number of agents $n\in\{1,3,5\}$. We repeat and report results for $3$ runs with different seeds. See Table~\ref{apx:tab:ppo} and Table~\ref{apx:tab:shac} for a complete list of hyperparameters.

@@

\paragraph{Dispersion}
Firstly, we compare the performance of these algorithms on a non-differentiable scenario (thus SHAC is not aplicable), dispersion. We show the comparison between \fname{} and PPO for increasing number of agents for the transformer architecture (see Fig.~\ref{apx:fig:experiments-mlp}). In single agent setting, show in Fig.~\ref{fig:dispersion-mlp-1}, both algorithm perform well, albeit \fname{} tends to converge after fewer episodes. With $3$ agents, shown in Fig.~\ref{fig:dispersion-transformer-3}, PPO fails to converge, meanwhile \fname{} converges in less than $10,000$ episodes. With $5$ agents, shown in Fig.~\ref{fig:dispersion-transformer-5}, PPO fails to converge, meanwhile \fname{} achieves high rewards without however triggering the early stopping. In this last scenario, \fname{} manifest also high reward variance.

\paragraph{Transport}
Now, we compare the performance of these algorithms on a fully differentiable scenario, transport. On the single agent setting, shown in Fig.~\ref{fig:transport-mlp-1}, PPO fails to converge, meanwhile both SHAC and \fname{} perform similarly well. While \fname{} appears to converge earlier SHAC appear to be more consistent. With $3$ and $5$ agents, shown in Fig.~\ref{fig:transport-transformer-3} and \ref{fig:transport-transformer-5}, PPO fails to converge, meanwhile both SHAC and \fname{} perform similarly well albeit SHAC is more stable. For the MLP only setting see Fig.~\ref{apx:fig:experiments-mlp}

\paragraph{Discovery}
Similar consideration hold also for the discovery scenario. In the single agent setting, shown in Fig.~\ref{fig:discovery-mlp-1}, \fname{} performs better, but comparitavely, to PPO . However, with $3$ and $5$ agents, shown in Fig.~\ref{fig:discovery-transformer-3} and \ref{fig:discovery-transformer-5}, PPO fails to converge, meanwhile \fname{} performs better. For the MLP only setting see Fig.~\ref{apx:fig:experiments-mlp}

>>>>

\begin{figure*}[!t]
    \includegraphics[width=\textwidth]{figs/main-transformer.pdf}
    \caption{Comparison between \fname{}, PPO, and SHAC for increasing number of agents for Dispersion, Transport, Discovery, and Sampling scenarios for the Transformer Architecture.}
    \label{fig:experiments}
\end{figure*}

@@

\paragraph{Scenarios}
As mentioned previously, we choose a few multi-agent scenarios from the VMAS\footnote{\url{https://github.com/proroklab/VectorizedMultiAgentSimulator}} package to test \fname{} aganist SHAC and PPO. In particular, we choose the following scenario:
\begin{compactitem}
    \item \textbf{Dispersion} (\Cref{fig:dispersion}): 
        \sentence{Environment} a scenario where $n$ agents need to reach $n$ goal locations, in such case the agent receives a sparse reward of $1$. Agents observe only their own position, velocity and relative position to each goal. 
        \sentence{Differentiability} Despite the transition function being differentiable, the reward function is not. Therefore, only PPO and \fname{} are applicable to this scenario. 
        \sentence{World Model} The world model has enough information to predict the next states accurately.
        \sentence{Max Reward} The maximum reward is $n$.
    \item \textbf{Transport} (\Cref{fig:transport}): 
        \sentence{Environment} a scenario where $n$ agents need to collaborate to push a package into a goal location. Agents observe their own position, velocity, relative position to the package and relative position between package and the goal. The agents receive a reward proportional to the distance between package and the goal. 
        \sentence{Differentiability} In this scenario, both the reward and transition functions are differentiable. 
        \sentence{World Model} The world model has enough information to predict the next states accurately.
        \sentence{Max Reward} The maximum reward varies from environement to environment. It is the scaled distance between the package and the goal times the number of agents.
    \item \textbf{Discovery} (\Cref{fig:discovery}): 
        \sentence{Environment} a scenario where $n$ agents need to collect $k$ goal place randomly. A goal is collected if at least $s$ agents stand in proximity to it, in this case they receive a reward of $1$. If agents collide with each other, they receive a penalty. Every agent observes its position, velocity, lidar measurements to other agents and goals. When $n=1$ we set $s=1$ and $k=7$. When $n>1$ we set $s=2$.
        \sentence{Differentiability} In this scenario, only the transition function is differentiable. Therefore, only PPO and \fname{} are applicable to this scenario. 
        \sentence{Max Reward} The maximum rewards depends on the number evaluation steps, but for early stopping purposes, we set it to $k$. 
        \sentence{World Model} Note that, a world model (or an action world model) that uses only the agents observation cannot predict the evolution correctly, as goals may not be visible to the agents. For this reason, this scenario is particularly challenging for \fname{} as we do not provide the full world state to the world model.
     \item \textbf{Sampling} (\Cref{fig:sampling}): 
        \sentence{Environment} a scenario where $n$ agents need to collect rewards from a grid. The reward are sampled from $k$ gaussians. Agents observe their own position, velocity, lidar values to each other, and rewards in a 3x3 grid around themselves.  
        \sentence{Differentiability} In this scenario, both the reward and transition functions are differentiable. 
        \sentence{Max Reward} The maximum reward depends on where the gaussians are placed.
        \sentence{World Model} Similarly to discovery, this scenario is particularly challenging for \fname{} as we do not provide the full world state, but only the agents observations, to the world model.
\end{compactitem}

@@

\paragraph{Hyperparameters}
For all networks we use learning rate of $1e\text{-}3$ with Adam Optimizer~\cite{Kingma14}. For all scenarios, each training episode consists of $512$ environments of $32$ steps each. Each validation episode is composed of $512$ environments of $512$ steps each. We employ early stopping when the agents achieve $90\%$ of the maximum reward in $90\%$ of the episode's environments. If early stopping is not triggered, we stop training after $20,000$ episodes. The discount factor is set to $0.99$ and lambda factor is set to $0.95$. We report results for increasing number of agents $n\in\{1,3,5\}$. We repeat and report results for $3$ runs with different seeds. See \Cref{apx:tab:ppo} and \Cref{apx:tab:shac} for a complete list of hyperparameters.

@@

\paragraph{Dispersion}
Firstly, we compare the performance of these algorithms on a non-differentiable scenario (thus SHAC is not aplicable), dispersion. We show the comparison between \fname{} and PPO for increasing number of agents for the transformer architecture (see \Cref{apx:fig:experiments-mlp}). In single agent setting, show in \Cref{fig:dispersion-mlp-1}, both algorithm perform well, albeit \fname{} tends to converge after fewer episodes. With $3$ agents, shown in \Cref{fig:dispersion-transformer-3}, PPO fails to converge, meanwhile \fname{} converges in less than $10,000$ episodes. With $5$ agents, shown in \Cref{fig:dispersion-transformer-5}, PPO fails to converge, meanwhile \fname{} achieves high rewards without however triggering the early stopping. In this last scenario, \fname{} manifest also high reward variance.

\paragraph{Transport}
Now, we compare the performance of these algorithms on a fully differentiable scenario, transport. On the single agent setting, shown in \Cref{fig:transport-mlp-1}, PPO fails to converge, meanwhile both SHAC and \fname{} perform similarly well. While \fname{} appears to converge earlier SHAC appear to be more consistent. With $3$ and $5$ agents, shown in \Cref{fig:transport-transformer-3} and \Cref{fig:transport-transformer-5}, PPO fails to converge, meanwhile both SHAC and \fname{} perform similarly well albeit SHAC is more stable. For the MLP only setting see \Cref{apx:fig:experiments-mlp}

\paragraph{Discovery}
Similar consideration hold also for the discovery scenario. In the single agent setting, shown in \Cref{fig:discovery-mlp-1}, \fname{} performs better, but comparitavely, to PPO . However, with $3$ and $5$ agents, shown in \Cref{fig:discovery-transformer-3} and \Cref{fig:discovery-transformer-5}, PPO fails to converge, meanwhile \fname{} performs better. For the MLP only setting see \Cref{apx:fig:experiments-mlp}

### 2025-01-19 paper/icml25/first/sects/related.tex
prefer Cref on ref

\paragraph{Differentiable Engines}
Game engines~\cite{Bellemare13,Balla23} and Physics engines \cite{Todorov12,Coumans16} have been widely used in the field of policy learning. Recently, there has been a trend towards differentiable engines, which allows for the use of gradient-based optimization methods, such as backpropagation. Notable examples are \cite{Freeman21,Howell22,Genesis24}. These frameworks may be built on top of auto differentiation libraries such \cite{Bradbury18,Ansel24} (this is the case of VMAS) or directly with analytical gradients \cite{Carpentier18,Werling21}. Further, several works tackle the challenge of the non-smoothness of the contact dynamics \cite{Degrave19,Moritz20}

@@

\paragraph{Reinforcement Learning on Differentiable Engines}
Precursors of SHAC with similar intuition are PODS~\cite{Mora21} and CE-APG~\cite{Gillen22}. In \cite{Lv23}, the authors introduce rendering to enhance model-based RL. \cite{Zheng24} employes a differentiable symbolic expression search engine. Trajectory optimization approach is taken in \cite{Wan24}. \cite{Georgiev24} takles unstable gradient due to stiff dynamics by adaptively truncating trajectories when these occur. With respect to this approach, we take a substantially different method, by emulating the differentiable environment with a trained neural network.

>>>>

\paragraph{Differentiable Engines}
Game engines~\cite{Bellemare13,Balla23} and Physics engines~\cite{Todorov12,Coumans16} have been widely used in the field of policy learning. Recently, there has been a trend towards differentiable engines, which allows for the use of gradient-based optimization methods, such as backpropagation. Notable examples are~\cite{Freeman21,Howell22,Genesis24}. These frameworks may be built on top of auto differentiation libraries such~\cite{Bradbury18,Ansel24} (this is the case of VMAS) or directly with analytical gradients~\cite{Carpentier18,Werling21}. Further, several works tackle the challenge of the non-smoothness of the contact dynamics~\cite{Degrave19,Moritz20}

@@

\paragraph{Reinforcement Learning on Differentiable Engines}
Precursors of SHAC with similar intuition are PODS~\cite{Mora21} and CE-APG~\cite{Gillen22}. In~\cite{Lv23}, the authors introduce rendering to enhance model-based RL. \cite{Zheng24} employes a differentiable symbolic expression search engine. Trajectory optimization approach is taken in~\cite{Wan24}. \cite{Georgiev24} takles unstable gradient due to stiff dynamics by adaptively truncating trajectories when these occur. With respect to this approach, we take a substantially different method, by emulating the differentiable environment with a trained neural network.
